---
title: "07 Gradient Boosting"
subtitle: "ML - Trees"
author: "Lasse Hyldig Hansen"
date: "2023-06-06"
output: 
  html_document:
    toc: true
---
## Load required packages

```{r}
install.packages("gbm")
library(gbm)
```


## Gradient boosting

Last, but not least, we move on to gradient boosting. Gradient boosting, our last topic, elegantly combines concepts from the previous methods. As a “boosting” method, gradient boosting involves iteratively building trees, aiming to improve upon misclassifications of the previous tree. Gradient boosting also borrows the concept of sub-sampling the variables (just like Random Forests), which can help to prevent overfitting.

While it is too much to express in this tutorial, the biggest innovation in gradient boosting is that it provides a unifying mathematical framework for boosting models. The approach explicitly casts the problem of building a tree as an optimization problem, defining mathematical functions for how well a tree is performing (which we had before) and how complex a tree is. In this light, one can actually treat AdaBoost as a “special case” of gradient boosting, where the loss function is chosen to be the exponential loss.

Let’s build a gradient boosting model.

## Building model

```{r}
# Set seed for reproducibility
set.seed(321)

# Build the gradient boosting model
mdl <- gbm(formula = as.character(actualhospitalmortality_enc) ~ age + acutephysiologyscore, 
           distribution = "bernoulli", 
           data = train_df, 
           n.trees = 15)
```

## Visualizing Decision Boundaries

```{r}
# Generate grid of points
grid <- expand.grid(age = seq(min(train_df$age), max(train_df$age), length.out = 100),
                    acutephysiologyscore = seq(min(train_df$acutephysiologyscore), max(train_df$acutephysiologyscore), length.out = 100))

# Predict class label for each point in the grid
grid$prediction <- predict(mdl, newdata = grid, type = "response")



# Create decision boundary plot
gradient_boosted_tree_plot <- ggplot(grid, aes(x = age, y = acutephysiologyscore, fill = as.factor(ifelse(prediction > 0.5, 1, 0)))) +
  geom_tile() +
  scale_fill_manual(values = c("blue", "red")) +
  labs(title = 'Gradient Boosted Tree (Final Decision Surface)',
       x = "Age",
       y = "Acute Physiology Score",
       fill = "Predicted Class") +
  theme_classic()

# Print the plot
print(gradient_boosted_tree_plot)
```

