---
title: "08 Performance"
subtitle: "ML - Trees"
author: "Lasse Hyldig Hansen"
date: "2023-06-06"
output: 
  html_document:
    toc: true
---

## packages

```{r}
# Load necessary libraries
library(rpart)
library(adabag)
library(gbm)
library(ipred)
library(pROC)

```

## Comparing model performance
We’ve now learned the basics of the various tree methods and have visualized most of them. Let’s finish by comparing the performance of our models on our held-out test data. Our goal, remember, is to predict whether or not a patient will survive their hospital stay using the patient’s age and acute physiology score computed on the first day of their ICU stay.

```{r}
# Create an empty list to store models
clf <- list()

# Train the Decision Tree model
clf[['Decision Tree']] <- rpart(actualhospitalmortality_enc ~ age + acutephysiologyscore, data = train_df, method = "class")

# Train the AdaBoost model
clf[['AdaBoost']] <- boosting(formula = actualhospitalmortality_enc ~ age + acutephysiologyscore, 
                data = train_df, 
                boos = TRUE, 
                mfinal = 6)

# Train the Gradient Boosting model
clf[['Gradient Boosting']] <- gbm(formula = as.factor(actualhospitalmortality_enc) ~ age + acutephysiologyscore, 
           distribution = "bernoulli", 
           data = train_df, 
           n.trees = 15)

# Train the Random Forest model
clf[['Random Forest']] <- randomForest(actualhospitalmortality_enc ~ age + acutephysiologyscore, data = train_df, ntree = 10)

# Train the Bagging model
clf[['Bagging']] <- bagging(actualhospitalmortality_enc ~ age + acutephysiologyscore, data = train_df, nbagg = 10)


## THIS NEEDS TO BE REVISED!

# Print AUROC for each model
for(curr_mdl in names(clf)) {
  # Make predictions on test data
  if (curr_mdl == 'AdaBoost') {
    yhat <- predict(clf[[curr_mdl]], newdata = test_df, type = 'response')$prob[,2]
  } else if (curr_mdl == 'Gradient Boosting') {
    yhat <- predict.gbm(clf[[curr_mdl]], newdata = test_df, n.trees = 15, type = 'response')
  } else {
    yhat <- predict(clf[[curr_mdl]], newdata = test_df, type = 'prob')[,2]
  }
  
  # Calculate AUROC
  roc_obj <- roc(test_df$actualhospitalmortality_enc, yhat)
  score <- auc(roc_obj)
  
  # Print AUROC score
  cat(sprintf('AUROC for %s: %f\n', curr_mdl, score))
}
```

