---
title: "04 - Other Plots"
subtitle: "Data Visualization"
author: "Renata Prôa"
date: "2023-07-19"
questions:
- "What are the other geometries that I can use to show my data?"
- "How can I deal with many continuous variables?"

objectives:
- "Explore different types of plots."
- "Learn to transform numerical variables into categorical variables."

keypoints:
- "Visualization sometimes can benefit from spliting countinuous data into groups."
- "Colors can improve visualization even when redundant."

output: 
  html_document:
    toc: true
---

#### Import and Prepare Data

```{r, message=FALSE, warning=FALSE}
cohort <- read_csv("../data/eicu_cohort.csv")

cohort <- cohort %>%
  mutate(age = as.integer(age)) %>%
  mutate(age = if_else(is.na(age), 91.5, age)) 

names(cohort)[names(cohort) == "unabridgedhosplos"] <- "length_of_stay"
names(cohort)[names(cohort) == "meanbp"] <- "mean_blood_pressure"
names(cohort)[names(cohort) == "wbc"] <- "white_cell_count"
#print(colnames(cohort))
```

## Separating Continuous Data Into Categories

Sometimes, visualization can benefit from splitting the cohort into groups based on ranges of a continuous variable. One of the reasons for that is that categorical variables can be shown as subplots. We have a finite number of dimensions that we can see, as humans we can only deal with 3 spacial dimensions (and in the computer we are actually limited to 2)! So in order to increase the dimensionality of our graphs we take advantage of other elements that we can vary such as colors, shapes and directions, including multiplication of the figure (subplots). But some of those elements can only vary in a discrete manner or at least are more easily differentiated when varied in a discrete manner. Subplots, i.e. dividing the plot in groups, is one very import resource for showing more information, but it requires that we have a finite number of possible values! Hence the utility of splitting continuous variables into groups.

For example, we can separate the cohort into age groups by decade.

```{r, message=FALSE}
cohort <- cohort %>%
  mutate(age_group = cut(age, breaks = seq(0, 100, by = 10), labels = c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80-89", "90-99"), right = FALSE))
head(cohort)
```

#### BMI Groups

```{r}
# Calculate BMI
cohort <- cohort %>% 
  mutate(height_m = admissionheight / 100,  # Convert height to meters
         bmi = admissionweight / (height_m^2))
# Split in groups
bmi_categories <- c("< 18.5", "18.5 - 24.9", "24.9 - 29.9", "> 29.9")
bmi_ranges <- c(0, 18.5, 24.9, 29.9, Inf)  # Ranges for each category
cohort <- cohort %>% 
  mutate(bmigroup = cut(bmi, breaks = bmi_ranges, labels = bmi_categories, 
                        include.lowest = TRUE))
```

#### Blood Pressure Groups

```{r}
bp_categories <- c("Normal", "Elevated", "Hypertension Stage 1", "Hypertension Stage 2")
cohort <- cohort %>%
  mutate(bp_category = cut(mean_blood_pressure, 
                           breaks = c(-Inf, 120, 130, 140, Inf), 
                           labels = bp_categories,
                           include.lowest = TRUE))
```

Check the new variables!

```{r}
head(cohort)
```

# Other Plots

#### Split Bar Plots

```{r, message=FALSE}
library(gridExtra)

p1 <- ggplot(data = cohort) +
      geom_bar(aes(x = actualhospitalmortality, fill = gender)) +
      labs(title = "Hospital Mortality", x = "Outcome", 
           y = "Frequency", fill = "Gender") +
      theme(plot.title = element_text(size = 11))

p2 <- ggplot(data = cohort) +
      geom_bar(aes(x = actualhospitalmortality, fill = gender), 
               position = "dodge") +
      labs(title = "Hospital Mortality", x = "Outcome", 
           y = "Frequency", fill = "Gender") +
      theme(plot.title = element_text(size = 11))

p3 <- ggplot(data = cohort) +
      geom_bar(aes(x = age_group, fill = gender), position = "dodge") +
      labs(title = "Hospital Mortality", x = "Outcome", 
           y = "Frequency", fill = "Gender") +
      theme(plot.title = element_text(size = 11))

grid.arrange(p1, p2, ncol = 2)
print(p3)
```

#### Box Plots and Violin Plots

Boxplots display the 25th percentile, median, and 75th percentile of a distribution. The whiskers (vertical lines) capture roughly 99% of a normal distribution, and observations outside this range are plotted as points representing outliers. Violin plots are similar to kernel density plots, but are mirrored and rotated 90 degrees.

```{r}
p <- ggplot(data = cohort, aes(x = actualhospitalmortality, y = heartrate)) +
     labs(title = "Heart Rate by Hospital Mortality", x = "Outcome", 
          y = "Heart Rate", fill = "Gender") +
     theme(plot.subtitle = element_text(face = "italic"))

p1 <- p + geom_boxplot() + labs(subtitle = "Boxplot")
p2 <- p + geom_boxplot(notch = TRUE, fill = "lightblue") + labs(subtitle = "Notched Boxplot")
p3 <- p + geom_violin(fill = "lightgreen") + labs(subtitle = "Violin Plot")
p4 <- p + geom_violin(fill = "cornflowerblue") + 
      geom_boxplot(fill = "orange", width = .2) +
      labs(subtitle = "Violin Plot with Boxplot")

print(p1)
print(p2)
print(p3)
print(p4)
```

#### Ridgeline plots

Ridgeline plots, or joyplots display the distribution of a quantitative variable for several groups. They’re similar to kernel density plots with vertical faceting, but take up less room.

```{r, message=FALSE}
#install.packages("ggridges")
library(ggridges)
p <- ggplot(data = cohort, 
            aes(x = creatinine, y = age_group, fill = age_group)) +
     geom_density_ridges() + 
     labs(title = "Creatinine by Age Group", x = "Creatinine", 
          y = "Age Group") +
     theme_ridges() +
     theme(legend.position = "none")

print(p)

```

Although we don't need to vary the color here, it makes it easier to visualize and it looks more aesthetically pleasing. Apply to the other variables:

```{r, message=FALSE, warning=FALSE}

variables <- c("admissionweight","length_of_stay","acutephysiologyscore", "apachescore","heartrate","mean_blood_pressure","creatinine","temperature","respiratoryrate","white_cell_count","admissionheight")

# Create a function to generate ridge plots
generate_ridge_plot <- function(variable) {
  ggplot(data = cohort, 
         aes(x = .data[[variable]], y = age_group, fill = age_group)) +
    geom_density_ridges() + 
    labs(title = paste(variable, "by Age Group"), x = variable, 
         y = "Age Group") +
    theme_ridges() +
    theme(legend.position = "none")
}

# Create and display ridge plots for each variable
plot_list <- lapply(variables, generate_ridge_plot)

# Print the ridge plots
for (p in plot_list) {
  print(p)
}

```

#### Pie Charts

```{r}
categorical_vars <- cohort %>% select_if(is.factor)
for (col in names(categorical_vars)) {
  pie_data <- categorical_vars %>%
    count(!!sym(col))
  
  pie_chart <- ggplot(pie_data, aes(x = "", y = n, fill = !!sym(col))) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    labs(title = paste("Pie Chart of", col))
  
  print(pie_chart)
}
```

#### Facets

"Faceting" provides an alternative way of grouping plots into several sub-plots. The easiest way to distribute a graph into multiple panels is provided by `facet_wrap()`.

```{r}
p <- ggplot(data = cohort) +
     geom_point(aes(age, apachescore, color = gender)) +
     scale_shape_manual(values=c(16, 3), labels = c("Alive", "Expired"), name = 'Hospital Mortality') +
     scale_color_manual(values = c("lightcoral", "mediumseagreen", "grey"), 
                        labels = c("Female", "Male", "Unknown"), name = "Gender") 
p1 <- p + facet_wrap(~actualhospitalmortality)
p2 <- p + facet_wrap(~bmigroup)
p3 <- p + facet_wrap(~bp_category)
print(p1)
print(p2)
print(p3)
```

#### Mean Apache Score by Age by Gender and Actual Mortality

```{r, message=FALSE}

mean_apache_by_age_by_gender <- aggregate (apachescore ~ age + gender, data = cohort, FUN = mean)
mean_apache_by_age_by_mortality <- aggregate (apachescore ~ age + actualhospitalmortality, 
                                              data = cohort, FUN = mean)

p1 <- ggplot(mean_apache_by_age_by_gender) + geom_smooth(aes(x = age, y = apachescore, color = gender)) +
      ggtitle("Mean APACHE Score by Age") + theme_economist()
p2 <- ggplot(mean_apache_by_age_by_mortality) + geom_smooth(aes(x = age, y = apachescore, color = actualhospitalmortality)) +
      ggtitle("Mean APACHE Score by Mortality") + theme_economist()

print(p1)
print(p2)
```


